Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Patterns in Australian Spatial Statistics

### 2025-07-09 paper/kobakian-cook-anzjs.qmd
tidied up speed section

```{r pdetection_group}
# Tidy for analysis
d <- d %>% 
  separate(image_name, c("nothing", "trend", "location", "type", "extra"), remove = FALSE) %>%
  select(-nothing, -extra) %>%
  mutate(location = as.numeric(location), 
    # detect measures the accuracy of the choice
         detect = ifelse(location == choice, 1, 0)) %>% 
  mutate(trend = case_when(
    trend == "nwse" ~ "NW-SE",
    trend == "cities" ~ "All Cities",
    trend == "three" ~ "Three Cities")) %>% 
  mutate(trend = fct_relevel(trend,"All Cities","Three Cities", "NW-SE")) %>% 
  mutate(type = case_when(
    type == "hex" ~"Hex.",
    TRUE~"Choro.")) %>% 
    mutate(detect_f = factor(detect, levels = c(0,1), labels = c("Detected? No", "Detected? Yes")))

@@

```{r}
#| label: fig-beeswarm
#| fig-cap: "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The coloured point indicates average time taken for each map type. Although some participants take just a few seconds per evaluation, and some take as much as 60 seconds, but there is very little difference in time taken between map types."
#| fig-height: 6
#| fig-width: 5
#| out-width: 80%
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75))
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=5, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_manual("", values = type_colors) +
  facet_grid(trend~detect_f) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="none")
```

@@

# Check time difference between default choice
d %>% mutate(zeros = if_else(choice == 0, 0, 1)) %>%
  group_by(zeros) %>%   
  summarise(m=median(time_taken), 
            mn=min(time_taken),
            mx=max(time_taken),
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75),
            n=n())

d %>% mutate(zeros = if_else(choice == 0, 0, 1)) %>%
  group_by(zeros, detect_f) %>%   
  summarise(m=median(time_taken), 
            mn=min(time_taken),
            mx=max(time_taken),
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75),
            n=n())

# Summarise across subjects
contributor_smry <- d %>% group_by(contributor) %>%   
  summarise(m_d=mean(detect), 
            s_d=sd(detect),
            m_t=mean(time_taken), 
            s_t=sd(time_taken)) 
ggplot(contributor_smry, aes(m_t, m_d)) +
  geom_point() +
  ylab("Detection rate") +
  xlab("Time taken") +
  xlim(c(0, 45)) +
  theme(aspect.ratio = 1)
ggplot(contributor_smry, aes(m_t, s_t)) +
  geom_point() +
  xlab("Average") +
  ylab("SD") +
  theme(aspect.ratio = 1)
ggplot(contributor_smry, aes(m_d, s_d)) +
  geom_point() +
  xlab("Average") +
  ylab("SD") +
  theme(aspect.ratio = 1)
```

@@

@fig-beeswarm shows horizontally jittered dot plots to contrast the time taken by participants to evaluate each lineup when viewing each type of display. The time are also separated by trend model and whether the data plot was detected or not detected. The time taken to complete an evaluation ranged from milliseconds to 60 seconds. The average time taken for type of display is shown as a large coloured dot on each plot. When considering the heights of the green and orange dots, there is little difference in the average time taken to read a choropleth or hexagon tile map. Comparing the same coloured dot across each trend model row, there is a slight increase in the time taken to correctly detected the data plot in the hexagon tile map lineup, but little difference in evaluation time for the choropleth display. However, there were substantially less correct detections for choropleth lineups for the Three cities and All Cities trends.

@@

 - Document creation: quarto [@Allaire_Quarto_2025], anzjs template [@quarto-anzjs], knitr [@knitr].
 - Lineup creation: nullabor [@nullabor], gstat [@gstat].
 - Data analysis: tidyverse [@tidyverse], ggthemes [@ggthemes], RColorBrewer [@RColorBrewer].
 - Plots: ggplot2 [@ggplot2], cowplot [@cowplot], png [@png], grid [@grid].
 - Modelling and summary presentation: lme4 [@lme4], kableExtra [@kableExtra].

>>>>

```{r pdetection_group}
# Tidy for analysis
d <- d %>% 
  separate(image_name, c("nothing", "trend", "location", "type", "extra"), remove = FALSE) %>%
  select(-nothing, -extra) %>%
  mutate(location = as.numeric(location), 
    # detect measures the accuracy of the choice
         detect = ifelse(location == choice, 1, 0)) %>% 
  mutate(trend = case_when(
    trend == "nwse" ~ "NW-SE",
    trend == "cities" ~ "All Cities",
    trend == "three" ~ "Three Cities")) %>% 
  mutate(trend = fct_relevel(trend,"All Cities","Three Cities", "NW-SE")) %>% 
  mutate(type = case_when(
    type == "hex" ~"Hex.",
    TRUE~"Choro.")) %>% 
    mutate(detect_f = factor(detect, levels = c(1, 0), labels = c("Detected? Yes", "Detected? No")))

@@

```{r}
#| label: fig-beeswarm
#| fig-cap: "The distribution of the time taken (seconds) to submit a response for each combination of trend, whether the data plot was detected, and type of display, shown using horizontally jittered dotplots. The coloured point indicates average time taken for each map type. Although some participants take just a few seconds per evaluation, and some take as much as 60 seconds, but there is very little difference in time taken between map types."
#| fig-height: 4
#| fig-width: 6
#| out-width: 100%
s <- d %>% group_by(type, trend, detect_f) %>%
  summarise(m=median(time_taken), 
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75)) 
ggplot() + 
  geom_quasirandom(data=d, aes(x=type, y=time_taken), alpha=0.9) + 
  #geom_hline(data=s, aes(yintercept=m, color=type)) +
  geom_point(data=s, aes(x=type, y=m, color=type), size=3, alpha=0.7) +
  #geom_errorbar(data=s, aes(x=type, ymin=q1, ymax=q3, color=type), width=0.3, size=2) +
  scale_color_manual("", values = type_colors) +
  facet_grid(detect_f~trend) + 
  ylab("Time taken (seconds)") + xlab("") +
  theme(legend.position="none")
```

@@

```{r}
# Check time difference between default choice
d_zeros <- d %>% mutate(zeros = if_else(choice == 0, 0, 1)) %>%
  group_by(zeros) %>%   
  summarise(m=median(time_taken), 
            mn=min(time_taken),
            mx=max(time_taken),
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75),
            n=n())

d_zeros_detect <- d %>% mutate(zeros = if_else(choice == 0, 0, 1)) %>%
  group_by(zeros, detect_f) %>%   
  summarise(m=median(time_taken), 
            mn=min(time_taken),
            mx=max(time_taken),
            q1=quantile(time_taken, 0.25), 
            q3=quantile(time_taken, 0.75),
            n=n())
# Summarise across subjects
contributor_smry <- d %>% group_by(contributor) %>%   
  summarise(m_d=mean(detect), 
            s_d=sd(detect),
            m_t=mean(time_taken), 
            s_t=sd(time_taken)) 
```

```{r}
#| eval: false
ggplot(contributor_smry, aes(m_t, m_d)) +
  geom_point() +
  ylab("Detection rate") +
  xlab("Time taken") +
  xlim(c(0, 45)) +
  theme(aspect.ratio = 1)
ggplot(contributor_smry, aes(m_t, s_t)) +
  geom_point() +
  xlab("Average") +
  ylab("SD") +
  theme(aspect.ratio = 1)
ggplot(contributor_smry, aes(m_d, s_d)) +
  geom_point() +
  xlab("Average") +
  ylab("SD") +
  theme(aspect.ratio = 1)
```

@@

@fig-beeswarm shows horizontally jittered dot plots to contrast the time taken by participants to evaluate each lineup facetted by map type and trend model. Each dot is an evaluation. The time taken to complete an evaluation ranged from fractions of a second to 60 seconds. The average time taken for type of display is shown as a large coloured dot on each plot, and show there is little difference in the average time taken to read a lineup made with either a choropleth or hexagon tile map.

That some evaluations occurred within milliseconds is a little surprising. Investigating whether this was related the `r d_zeros$n[1]` of `r d_zeros$n[1]+d_zeros$n[2]` evaluations where participants left the default choice of 0, and we find it is not; most of these people took the routine time to examine, and then left it at the default suggesting that they just could not pick one as different. This is the same as a non-detect. On a per participant basis, the average time per lineup over the 12 evaluations ranged between `r round(min(contributor_smry$m_t), 2)` and `r round(max(contributor_smry$m_t), 2)` seconds. The correlation between average detection rate and time taken across subjects `r round(cor(contributor_smry$m_d, contributor_smry$m_t), 2)` which is weakly positive. This is similar to what we have found in other studies, some subjects are especially fast and accurate in visual evaluation, and conversely some subjects are quite slow but inaccurate.

@@

 - Document creation: quarto [@Allaire_Quarto_2025], anzjs template [@quarto-anzjs], knitr [@knitr].
 - Lineup creation: nullabor [@nullabor], gstat [@gstat].
 - Data analysis: tidyverse [@tidyverse], ggthemes [@ggthemes], RColorBrewer [@RColorBrewer].
 - Plots: ggplot2 [@ggplot2], cowplot [@cowplot], png [@png], grid [@grid], ggbeeswarm [@ggbeeswarm], ggmosaic [@ggmosaic].
 - Modelling and summary presentation: lme4 [@lme4], kableExtra [@kableExtra].

### 2025-07-09 paper/kobakian-cook-anzjs.tex
tidied up speed section

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{kobakian-cook-anzjs_files/figure-pdf/fig-beeswarm-1.pdf}

@@

Figure~\ref{fig-beeswarm} shows horizontally jittered dot plots to
contrast the time taken by participants to evaluate each lineup when
viewing each type of display. The time are also separated by trend model
and whether the data plot was detected or not detected. The time taken
to complete an evaluation ranged from milliseconds to 60 seconds. The
average time taken for type of display is shown as a large coloured dot
on each plot. When considering the heights of the green and orange dots,
there is little difference in the average time taken to read a
choropleth or hexagon tile map. Comparing the same coloured dot across
each trend model row, there is a slight increase in the time taken to
correctly detected the data plot in the hexagon tile map lineup, but
little difference in evaluation time for the choropleth display.
However, there were substantially less correct detections for choropleth
lineups for the Three cities and All Cities trends.

@@

\begin{itemize}
\tightlist
\item
  Document creation: quarto \citep{Allaire_Quarto_2025}, anzjs template
  \citep{quarto-anzjs}, knitr \citep{knitr}.
\item
  Lineup creation: nullabor \citep{nullabor}, gstat \citep{gstat}.
\item
  Data analysis: tidyverse \citep{tidyverse}, ggthemes \citep{ggthemes},
  RColorBrewer \citep{RColorBrewer}.
\item
  Plots: ggplot2 \citep{ggplot2}, cowplot \citep{cowplot}, png
  \citep{png}, grid \citep{grid}.
\item
  Modelling and summary presentation: lme4 \citep{lme4}, kableExtra
  \citep{kableExtra}.
\end{itemize}

>>>>

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{kobakian-cook-anzjs_files/figure-pdf/fig-beeswarm-1.pdf}

@@

Figure~\ref{fig-beeswarm} shows horizontally jittered dot plots to
contrast the time taken by participants to evaluate each lineup facetted
by map type and trend model. Each dot is an evaluation. The time taken
to complete an evaluation ranged from fractions of a second to 60
seconds. The average time taken for type of display is shown as a large
coloured dot on each plot, and show there is little difference in the
average time taken to read a lineup made with either a choropleth or
hexagon tile map.

That some evaluations occurred within milliseconds is a little
surprising. Investigating whether this was related the 28 of 1080
evaluations where participants left the default choice of 0, and we find
it is not; most of these people took the routine time to examine, and
then left it at the default suggesting that they just could not pick one
as different. This is the same as a non-detect. On a per participant
basis, the average time per lineup over the 12 evaluations ranged
between 3.91 and 40.42 seconds. The correlation between average
detection rate and time taken across subjects 0.3 which is weakly
positive. This is similar to what we have found in other studies, some
subjects are especially fast and accurate in visual evaluation, and
conversely some subjects are quite slow but inaccurate.

@@

\begin{itemize}
\tightlist
\item
  Document creation: quarto \citep{Allaire_Quarto_2025}, anzjs template
  \citep{quarto-anzjs}, knitr \citep{knitr}.
\item
  Lineup creation: nullabor \citep{nullabor}, gstat \citep{gstat}.
\item
  Data analysis: tidyverse \citep{tidyverse}, ggthemes \citep{ggthemes},
  RColorBrewer \citep{RColorBrewer}.
\item
  Plots: ggplot2 \citep{ggplot2}, cowplot \citep{cowplot}, png
  \citep{png}, grid \citep{grid}, ggbeeswarm \citep{ggbeeswarm},
  ggmosaic \citep{ggmosaic}.
\item
  Modelling and summary presentation: lme4 \citep{lme4}, kableExtra
  \citep{kableExtra}.
\end{itemize}
